---
title: "Classwork 10 - Tidy Data"
subtitle: "Data Analytics 200 <br><br>" 
author: 
  - name: Ann Brennan
    affiliations:
      - name: SUNY Geneseo
  - <br>
date: "`r Sys.Date()`"

format: 
  html:
    toc: true
    number-sections: true
    code-fold: false       # https://quarto.org/docs/output-formats/html-code.html
    code-summary: "Show the code"
    code-tools: true
    highlight-style: espresso # atom-one tango espresso

execute: 
  echo: true
  eval: true
  warning: false
  message: false
  fig-width: 9
---

## Question 1

```{r}
library(tidyverse)
library(skimr)

billboard <- read_csv('https://bcdanl.github.io/data/billboard.csv')
```

## Q1a

-   Describe how the distribution of rating varies across week 1, week 2, and week 3 using the **faceted** histogram.

```{r}
q1a <- billboard %>% 
  pivot_longer(cols = wk1:wk76,
               names_to = "week",
               values_to = "rating") %>% 
  filter(week %in% c('wk1', 'wk2', 'wk3'))
ggplot(q1a,
       aes(x = rating)) +
  geom_histogram() +
  facet_wrap(.~ week)
```

## Q1b

-   Which artist(s) have the most number of tracks in `billboard` data.frame?

-   Do not double-count an artist's tracks if they appear in multiple weeks.

```{r}

```

## Question 2

```{r}
ny_pincp <- read_csv('https://bcdanl.github.io/data/NY_pinc_wide.csv')
```

## Q2a

-   Make `ny_pincp` longer.

```{r}
q2a <- ny_pincp %>% 
  pivot_longer(cols = pincp1969:pincp2019,
               names_to = "year",
               values_to = "pincp")
```

## Q2b

-   Provide both (1) ggplot code and (2) a simple comment to describe how overall the yearly trend of NY counties' average personal incomes are.

```{r}

```

## Question 3

```{r}
covid <- read_csv('https://bcdanl.github.io/data/covid19_cases.csv')
```

## Q3a

-   Keep only the following three variables, `date`, `countriesAndTerritories`, and `cases`.

-   Then make a wide-form data.frame of `covid` whose variable names are from `countriesAndTerritories` and values are from `cases`.

-   Then drop the variable `date`.

```{r}

```

## Q3b

-   Use the wide-form data.frame of `covid` to find the top 10 countries in terms of the correlation between their cases and the `USA` case.

    -   Use `cor(data.frame)`, which returns a matrix.

    -   Then convert it to data.frame using `as.data.frame(matrix)`

        ```{r}

        ```
