---
title: "Exercise Patterns Across Income and Education Levels in the United States"
subtitle: "DANL 310: Data Visualization and Presentation"
author:
  - name: Ann Brennan
    affiliation: SUNY Geneseo
date: last-modified

execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-width: 9
  fig-height: 6

format:
  html:
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-expand: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: atom-one
---

```{r setup}
#| include: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(sf)
library(skimr)
library(hrbrthemes)
library(rmarkdown)
library(maps)

theme_set(
  theme_ipsum() +
    theme(
      strip.background = element_rect(fill = "lightgray"),
      axis.title.x = element_text(size = rel(1.33)),
      axis.title.y = element_text(size = rel(1.33))
    )
)

```

# Project Overview

Physical activity is an important part of overall health. However, not everyone exercises the same amount, which may be influenced by social and economic factors. This topic highlights how health behaviors are shaped by more than just personal preference. Factors such as income and education can affect an individual’s access to time, resources, and opportunities for physical activity. This may contribute to differences in exercise rates across groups. This project examines how exercise participation varies across income and education levels in the United States. The main question is: **Are individuals with higher income and education more likely to engage in physical activity?**


# Data

```{r data-import}

brfss <- read.csv('C:/Users/15856/Downloads/Behavioral_Risk_Factor_Surveillance_System_(BRFSS)_Prevalence_Data_(2011_to_present)_20260428.csv')

paged_table(brfss)
```

The dataset used in this project is from the Behavioral Risk Factor Surveillance System (BRFSS), a national survey conducted by the Centers for Disease Control and Prevention (CDC). This survey collects self-reported information from adults in all 50 states on health behaviors and related factors. The data used in this project is from 2024. The unit of observation is a group defined by demographic categories, including income and education level. This analysis focuses on three main variables: exercise participation, income level, and education level. Exercise participation is measured as the percentage of individuals within each group who have engaged in physical activity in the past 30 days. Income and education are reported in categories. Although the data is self-reported, it still offers useful insights into health behavior across the United States. 


# Data Preparation and Transformation

To prepare for the analysis, the dataset was first filtered to include only respondents who had exercised in the past 30 days. It was also filtered to include the relevant demographic categories, including income and education. Observations with missing values were removed. Key variables such as exercise rates and sample sizes were converted to numeric format. The data was then grouped by income and education categories. A weighted average of exercise participation was calculated for each group using the sample size as weights. This ensures that groups with larger sample sizes have a greater influence on the results. 

```{r data-preparation}
brfss_clean <- brfss |>
  filter(Response == "Yes") |>
  filter(Break_Out_Category %in% c("Household Income", "Education Attained"))
  
brfss_clean <- brfss_clean |>
  filter(!is.na(Data_value), !is.na(Sample_Size))

brfss_clean <- brfss_clean |>
  mutate(
    Data_value = as.numeric(Data_value),
    Sample_Size = as.numeric(gsub(",", "", Sample_Size))
  )

income_df <- brfss_clean |>
  filter(Break_Out_Category == "Household Income") |>
  group_by(Break_Out) |>
  summarize(
    exercise_rate = weighted.mean(Data_value, Sample_Size, na.rm = TRUE)
  ) |>
  rename(income = Break_Out)

income_df$income <- factor(income_df$income, levels = c(
  "Less than $15,000",
  "$15,000-$24,999",
  "$25,000-$34,999",
  "$35,000-$49,999",
  "$50,000-$99,999",
  "$100,000-$199,999",
  "$200,000+"
))

income_df <- income_df |>
  filter(!is.na(income))

education_df <- brfss_clean |>
  filter(Break_Out_Category == "Education Attained") |>
  group_by(Break_Out) |>
  summarize(
    exercise_rate = weighted.mean(Data_value, Sample_Size, na.rm = TRUE)
  ) |>
  rename(education = Break_Out)

education_df$education <- factor(education_df$education, levels = c(
  "Less than H.S.",
  "H.S. or G.E.D.",
  "Some post-H.S.",
  "College graduate"
))

```


# Descriptive Statistics

## Basic Summary

```{r summary-statistics}
skim(brfss_clean)
```

## Important Statistics

Exercise rate:
```{r exercise-rate}
# overall exercise rate
overall_exercise <- weighted.mean(
  brfss_clean$Data_value,
  brfss_clean$Sample_Size,
  na.rm = TRUE
)

overall_exercise

```

Income groups range:
```{r income}
range(income_df$exercise_rate, na.rm = TRUE)
```

Education groups range:
```{r education}
range(education_df$exercise_rate, na.rm = TRUE)
```

The descriptive statistics show clear variation across socioeconomic groups. The overall average exercise rate is about 78%. Income groups range from about 62% to 92%, whereas education groups range from about 59% to 87%. The differences within groups are slightly more pronounced across income levels, suggesting larger gaps by income. 


# Data Visualization

## Visualization 1: Exercise Rates by Income

```{r visualization-1}
ggplot(income_df, aes(x = income, y = exercise_rate)) +
  geom_col(fill = "#4E79A7") +
  labs(
    title = "Exercise Rates by Income",
    x = "Income Level",
    y = "Exercise Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

This chart shows a clear positive relationship between income and exercise rates. As income increases, the percentage of individuals who report exercising in the past 30 days also increases. This pattern suggests that higher income individuals are more likely to participate in physical activity. This could be because having a higher income provides greater access to resources that support exercise, such as gym memberships, safe neighborhoods, and flexible time. These resources may be less available to lower income groups. 

## Visualization 2: Exercise Rates by Education

```{r visualization-2}
ggplot(education_df, aes(x = education, y = exercise_rate)) +
  geom_col(fill = "#59A14F") +
  labs(
    title = "Exercise Rates by Education",
    x = "Education Level",
    y = "Exercise Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

This chart shows a similar positive relationship between education level and exercise rates. Individuals with less than a high school edcuation have the lowest exercise rate (about 60%), whereas college graduates have the highest exercise rate (about 90%). Each step up in education level is correlated with an increase in exercise participation. This suggests that education is strongly associated with healthier behaviors like exercise. Higher levels of education may increase awareness of the benefits of physical activity and correlate with occupations that give an individual more time or opportunity to exercise. 


```{r prepare data for visual 3}
#| include: FALSE

state_df <- brfss_clean |>
  group_by(Locationabbr) |>
  summarize(
    exercise_rate = weighted.mean(Data_value, Sample_Size, na.rm = TRUE)
  )

us_map <- map_data("state")

state_df <- state_df %>%
  mutate(region = tolower(state.name[match(Locationabbr, state.abb)]))

map_df <- left_join(us_map, state_df, by = "region")

```

## Visualization 3: Exercise Rates Across United States

```{r visualization-3}
ggplot(map_df, aes(long, lat, group = group, fill = exercise_rate)) +
  geom_polygon(color = "white", size = 0.2) +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#DEEBF7", high = "#3182BD", na.value = "grey90") +
  labs(
    title = "Exercise Rates Across United States",
    fill = "Exercise Rate (%)"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "right"
  )
```

This map shows geographic variation in exercise rates across states. Some states show much higher levels of physical activity than others. Western and a few northeastern states tend to have higher exercise rates (darker shades), while some southern and midwestern states appear to have lower rates (lighter shades). These regional differences may reflect socioeconomic and environmental factors, such as differences in income distribution, education levels, climate, and infrastructure. 


## Shiny App

```{=html}
<div style="display: flex; justify-content: center; margin: 1em 0;">
  <iframe
    src="http://127.0.0.1:6249"
    width="800"
    height="800"
    frameborder="0"
    style="border: none;"
  ></iframe>
</div>
```

Briefly describe the Shiny app and dashboard that accompany this project.


## Dashboard

- **Dashboard:** [Insert link here](YOUR_DASHBOARD_URL)

Explain how dashboard help users interact with or explore the project data.




## Connecting the Dots

Bring the visual evidence together.

Explain how your descriptive statistics, transformations, and visualizations work together to answer your guiding question or support your story theme.

You may discuss:

- the strongest pattern in the data,
- a surprising result,
- differences across groups,
- changes over time,
- geographic or categorical variation,
- practical, business, economic, social, cultural, or policy relevance,
- why your audience should care.

# Discussion

Interpret the meaning of your findings in plain language.

This project does **not** require regression, machine learning, or causal inference. Avoid claiming that one variable causes another unless your project design truly supports that claim.

Use careful language such as:

- The data suggest that...
- The pattern is consistent with...
- One possible explanation is...
- This comparison shows that...
- This trend raises questions about...

## Limitations

Discuss what your data story can and cannot show.

Possible limitations include:

- missing data,
- limited time period,
- measurement issues,
- small sample size,
- data not representing all populations,
- descriptive analysis cannot prove causality.

# Conclusion

Summarize your project in one or two paragraphs.

Return to your Big Idea and explain the main takeaway your audience should remember.

A strong conclusion should answer:

- What did the data reveal?
- Why does it matter?
- What should the audience remember after leaving your webpage?

# Future Directions

Briefly explain what you would explore next.

Examples:

- additional variables,
- a longer time period,
- different geographic areas,
- more detailed group comparisons,
- a dashboard or interactive visualization improvement,
- a follow-up analysis using modeling, if appropriate.

# References

List all sources used in your project.

Include:

- data sources,
- websites,
- reports,
- articles,
- R packages or tools, if relevant.

For online sources, include the full URL and date accessed.


