---
title: DANL Project
subtitle: "Climate Finance"
author: 
  - Ann Brennan
  - Alex Ventresca
  - Drew Cox

toc: true
toc-title: "Table of Contents"
toc-depth: 2
number-sections: true

fig-width: 9

execute:
  code-tools: true
  message: false
  warning: false

from: markdown+emoji
---

```{r setup}
#| include: false
library(tidyverse)
library(skimr)
library(ggthemes)
library(hrbrthemes)

theme_set(theme_ipsum()+
          theme(strip.background =element_rect(fill="lightgray"),
                axis.title.x = 
                  element_text(angle = 0,
                               size = rel(1.33),
                               margin = margin(10,0,0,0)),
                axis.title.y = 
                  element_text(angle = 0,
                               size = rel(1.33),
                               margin = margin(0,10,0,0))
                )
          )
```

# Introduction

About this project :clap:

# Data

```{r}
path <- 'https://bcdanl.github.io/data/climate_finance_energy.csv'
climate_finance <- read_csv(path)
```

## Variable Description

-   **Party:** a party (country) that provides a funding contribution to recipient country/region for their cliamte change project.

-   **Recipient country/region:** Recipient country or region

-   **Project/programme/activity:** Details in the climate change project

-   **Type of support:**

    -   adaptation if the climate change project is related to adaptation project.

    -   mitigation if the climate change project is related to mitigation project.

-   **Year:** Year that funding contribution is committed or provided.

-   **Contribution:** An amount of funding contribution for the climate change project (in USD).

-   **Status:**

    -   committed if a party commits to providing the funding contribution for the climate change project, but the funding contribution is NOT actually provided.

    -   provided if the funding contribution was provided for the climate change project.

-   **Energy:**

    -   TRUE if the project is energy-related;

    -   FALSE otherwise.

## Summary Statistics

```{r}
mpg <- ggplot2::mpg
```

```{r}
#| results: asis
#| echo: false
rmarkdown::paged_table(mpg) 
```

```{r}
skim(mpg) %>% 
  select(-n_missing)
```

## MPG and a Type of Cars

The following boxplot shows how the distribution of highway MPG (`hwy`) varies by a type of cars (`class`) :blue_car: :truck: :minibus:.

```{r}
ggplot(data = mpg) +
  geom_boxplot(aes(x = class, y = hwy, fill = class),
               show.legend = F) +
  labs(x = "Class", y = "Highway\nMPG") 
```
